---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

# Force English R messages while rendering, so the published README
# stays consistent regardless of the contributor's locale. Wrapped in
# tryCatch because some platforms reject "LC_MESSAGES".
invisible(tryCatch(
  Sys.setlocale("LC_MESSAGES", "C"),
  error = function(e) NULL,
  warning = function(w) NULL
))
Sys.setenv(LANGUAGE = "en")
```

# genproc

<!-- badges: start -->
[![R-CMD-check](https://github.com/danielrak/genproc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/danielrak/genproc/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/danielrak/genproc/graph/badge.svg)](https://app.codecov.io/gh/danielrak/genproc)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

`genproc` turns one-off iterative R procedures (`for`, `lapply()`,
`purrr::pmap()`, ...) into production-grade workflows by wrapping
them with orthogonal, composable execution layers. The goal is to
make the transition from *"a script that works on one case"* to
*"a system that runs reliably on many cases"* an architectural
step, not an improvised rewrite.

Two layers are always active and cannot be disabled:

- **Logged** — each case produces a structured log row with the
  real traceback (captured via `withCallingHandlers()`) and per-case
  timing.
- **Reproducibility** — each run records the R version, loaded
  package versions, execution environment, the exact iteration
  mask, and the spec of any optional layer used, so that two runs
  can be compared and any change in execution parameters is
  auditable.

Two optional layers can be composed with the defaults, at the
caller's choice:

- **Parallel** execution via the `future` ecosystem
  (`future.apply::future_lapply()`).
- **Non-blocking** execution: `genproc()` returns immediately with
  a `genproc_result` of status `"running"` while the run continues
  in a background future. Poll with `status()`, block with
  `await()`.

`genproc` has **zero Shiny dependency**. A companion package
(`genprocShiny`) will later build a UI on top of these functions.

## Installation

```r
# install.packages("remotes")
remotes::install_github("danielrak/genproc")
```

## A minimal example

A toy file-conversion task: read a few CSVs from one directory,
save them as RDS into another. One case per file.

```{r example-setup}
library(genproc)

# Synthetic workspace
src_dir <- file.path(tempdir(), "src")
dst_dir <- file.path(tempdir(), "dst")
dir.create(src_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(dst_dir, showWarnings = FALSE, recursive = TRUE)

write.csv(head(iris),       file.path(src_dir, "a.csv"), row.names = FALSE)
write.csv(head(mtcars),     file.path(src_dir, "b.csv"), row.names = FALSE)
write.csv(head(airquality), file.path(src_dir, "c.csv"), row.names = FALSE)

# Per-case function
convert <- function(src_dir, src_file, dst_dir, dst_file) {
  df <- read.csv(file.path(src_dir, src_file))
  saveRDS(df, file.path(dst_dir, dst_file))
}

# Iteration mask: one row per conversion
mask <- data.frame(
  src_dir  = src_dir,
  src_file = c("a.csv", "b.csv", "c.csv"),
  dst_dir  = dst_dir,
  dst_file = c("a.rds", "b.rds", "c.rds"),
  stringsAsFactors = FALSE
)

result <- genproc(convert, mask)
```

Every run returns a `genproc_result` — a structured list with a
stable shape across runs:

```{r example-summary}
result
```

The `log` data.frame holds one row per case, with `case_id`, the
mask parameter values (`src_dir`, `src_file`, `dst_dir`, `dst_file`
here), then `success`, `error_message`, `traceback`, and
`duration_secs`. Below we display a subset of those columns for
readability — `error_message` and `traceback` are `NA` on this
happy path:

```{r example-log}
result$log[, c("case_id", "src_file", "dst_file",
               "success", "duration_secs")]
```

If a case fails, the run continues — the error is captured, not
thrown. Below we point one row of the mask to a file that does not
exist; `b.csv` itself is left untouched on disk so subsequent
sections of this README can still reference it:

```{r example-error}
mask_with_missing <- mask
mask_with_missing$src_file[2] <- "does_not_exist.csv"

result2 <- genproc(convert, mask_with_missing)
result2$log[!result2$log$success,
            c("case_id", "src_file", "error_message")]
result2$n_success
result2$n_error
```

The traceback column holds the real R call stack of the failing
case (filtered for the `tryCatch`/`withCallingHandlers` machinery),
so debugging a failed case reads like a normal R error.

## The reproducibility snapshot

`result$reproducibility` is a plain list recording the R version,
OS, timezone, loaded package versions, the exact mask used, the
specs of any optional layer, and a fingerprint of every input file
referenced by the mask. This snapshot lives inside the result —
no side file to keep in sync:

```{r repro}
str(result$reproducibility, max.level = 1)
```

A taste of what is captured (first few package versions):

```{r repro-packages}
head(result$reproducibility$packages, 5)
```

## Detecting silent input drift

`result$reproducibility$inputs` records the size and mtime of every
input file the mask refers to. The intent is to flag the most
common reproducibility failure: re-running the same code on the
same paths after an upstream file has been silently rewritten.

genproc detects input columns automatically: any character column
of the mask whose values are existing files is treated as such.
Pass `track_inputs = FALSE` to skip the capture, or override the
heuristic with `input_cols = c(...)` (force) or
`skip_input_cols = c(...)` (exclude).

```{r inputs-setup}
mask_paths <- data.frame(
  csv_in = file.path(src_dir, c("a.csv", "b.csv", "c.csv")),
  stringsAsFactors = FALSE
)
do_one <- function(csv_in) nrow(read.csv(csv_in))

run0 <- genproc(do_one, mask_paths)
run0$reproducibility$inputs$files
```

`diff_inputs()` compares two runs and tells you which referenced
files have changed since the first one:

```{r inputs-drift}
# Rewrite a.csv with strictly more content (size changes)
write.csv(iris, file.path(src_dir, "a.csv"), row.names = FALSE)

run1 <- genproc(do_one, mask_paths)
diff_inputs(run0, run1)
```

The default method is `"stat"` (size + mtime). It detects every
legitimate modification at near-zero cost; a content-hash variant
is on the roadmap for stronger guarantees on adversarial workloads.

## Parallel execution

Dispatch cases across workers by passing a `parallel_spec()`:

```r
# Four workers, a temporary multisession plan, restored on exit
result <- genproc(
  convert, mask,
  parallel = parallel_spec(workers = 4)
)
```

The two mandatory layers remain active in parallel mode. Case order
is preserved in the log regardless of the order in which workers
return. Under the hood, `parallel_spec()` builds a configuration
object consumed by `future.apply::future_lapply()`.

If you manage `future::plan()` yourself (recommended across several
calls to amortize worker startup), pass `parallel_spec()` without a
strategy and your current plan is used unchanged.

## Non-blocking execution

Return immediately, keep the console, collect later:

```r
job <- genproc(
  convert, mask,
  nonblocking = nonblocking_spec()
)

status(job)        # "running" or "done"
job <- await(job)  # blocks until resolution
job$log
```

`nonblocking_spec()` can be composed with `parallel_spec()` — the
non-blocking wrapper envelops the parallel dispatch:

```r
job <- genproc(
  convert, mask,
  parallel    = parallel_spec(workers = 4),
  nonblocking = nonblocking_spec()
)
```

## Status

Lifecycle: **experimental**. The four execution layers (logged,
reproducibility, parallel, non-blocking) and the `genproc_result`
contract are committed to forward compatibility across the 0.x
series — existing fields are guaranteed not to be removed or
renamed. New fields and new optional layers may be added.

The 0.1.0 release is the first public submission and is not yet on
CRAN. Install from GitHub for now (see above).

## Learn more

See `vignette("genproc")` for a deep dive: anatomy of the log and
reproducibility snapshot, per-case error handling, composition
patterns, and roadmap.

## License

MIT. See [LICENSE.md](https://github.com/danielrak/genproc/blob/master/LICENSE.md).
