---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# genproc

<!-- badges: start -->
[![R-CMD-check](https://github.com/danielrak/genproc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/danielrak/genproc/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/danielrak/genproc/graph/badge.svg)](https://app.codecov.io/gh/danielrak/genproc)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

`genproc` turns one-off iterative R procedures (`for`, `lapply()`,
`purrr::pmap()`, ...) into production-grade workflows by wrapping
them with orthogonal, composable execution layers. The goal is to
make the transition from *"a script that works on one case"* to
*"a system that runs reliably on many cases"* an architectural
step, not an improvised rewrite.

Two layers are always active and cannot be disabled:

- **Logged** — each case produces a structured log row with the
  real traceback (captured via `withCallingHandlers()`) and per-case
  timing.
- **Reproducibility** — each run records the R version, loaded
  package versions, execution environment, the exact iteration
  mask, and the spec of any optional layer used, so that two runs
  can be compared and any parameter drift is auditable.

Two optional layers can be composed with the defaults, at the
caller's choice:

- **Parallel** execution via the `future` ecosystem
  (`future.apply::future_lapply()`).
- **Non-blocking** execution: `genproc()` returns immediately with
  a `genproc_result` of status `"running"` while the run continues
  in a background future. Poll with `status()`, block with
  `await()`.

`genproc` has **zero Shiny dependency**. A companion package
(`genprocShiny`) will later build a UI on top of these functions.

## Installation

```r
# install.packages("remotes")
remotes::install_github("danielrak/genproc")
```

## A minimal example

A toy file-conversion task: read a few CSVs from one directory,
save them as RDS into another. One case per file.

```{r example-setup}
library(genproc)

# Synthetic workspace
src_dir <- file.path(tempdir(), "src")
dst_dir <- file.path(tempdir(), "dst")
dir.create(src_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(dst_dir, showWarnings = FALSE, recursive = TRUE)

write.csv(head(iris),       file.path(src_dir, "a.csv"), row.names = FALSE)
write.csv(head(mtcars),     file.path(src_dir, "b.csv"), row.names = FALSE)
write.csv(head(airquality), file.path(src_dir, "c.csv"), row.names = FALSE)

# Per-case function
convert <- function(src_dir, src_file, dst_dir, dst_file) {
  df <- read.csv(file.path(src_dir, src_file))
  saveRDS(df, file.path(dst_dir, dst_file))
}

# Iteration mask: one row per conversion
mask <- data.frame(
  src_dir  = src_dir,
  src_file = c("a.csv", "b.csv", "c.csv"),
  dst_dir  = dst_dir,
  dst_file = c("a.rds", "b.rds", "c.rds"),
  stringsAsFactors = FALSE
)

result <- genproc(convert, mask)
```

Every run returns a `genproc_result`. The log contains one row per
case, with stable `case_id`, the parameter values, `success`,
`error_message`, `traceback`, and `duration_secs`:

```{r example-log}
result$log[, c("case_id", "src_file", "dst_file",
               "success", "duration_secs")]
```

If a case fails, the run continues — the error is captured, not
thrown. Here we delete one source file on purpose before a second
run:

```{r example-error}
file.remove(file.path(src_dir, "b.csv"))

result2 <- genproc(convert, mask)
result2$log[result2$log$success == FALSE,
            c("case_id", "src_file", "error_message")]
result2$n_success
result2$n_error
```

The traceback column holds the real R call stack of the failing
case (filtered for the `tryCatch`/`withCallingHandlers` machinery),
so debugging a failed case reads like a normal R error.

## The reproducibility snapshot

`result$reproducibility` is a plain list recording the R version,
OS, timezone, loaded package versions, the exact mask used, and
the specs of any optional layer. This snapshot lives inside the
result — no side file to keep in sync:

```{r repro}
str(result$reproducibility, max.level = 1)
```

## Parallel execution

Dispatch cases across workers by passing a `parallel_spec()`:

```r
# Four workers, a temporary multisession plan, restored on exit
result <- genproc(
  convert, mask,
  parallel = parallel_spec(workers = 4)
)
```

The two mandatory layers remain active in parallel mode. Case order
is preserved in the log regardless of the order in which workers
return. Under the hood, `parallel_spec()` builds a configuration
object consumed by `future.apply::future_lapply()`.

If you manage `future::plan()` yourself (recommended across several
calls to amortize worker startup), pass `parallel_spec()` without a
strategy and your current plan is used unchanged.

## Non-blocking execution

Return immediately, keep the console, collect later:

```r
job <- genproc(
  convert, mask,
  nonblocking = nonblocking_spec()
)

status(job)        # "running" or "done"
job <- await(job)  # blocks until resolution
job$log
```

`nonblocking_spec()` can be composed with `parallel_spec()` — the
non-blocking wrapper envelops the parallel dispatch:

```r
job <- genproc(
  convert, mask,
  parallel    = parallel_spec(workers = 4),
  nonblocking = nonblocking_spec()
)
```

## Status

Early development. The public API is not stable yet. The package
is not on CRAN.

## Learn more

See `vignette("genproc")` for a deep dive: anatomy of the log and
reproducibility snapshot, per-case error handling, composition
patterns, and roadmap.

## License

MIT. See [LICENSE.md](LICENSE.md).
