---
title: "genproc: deep dive"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{genproc: deep dive}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

The `README` introduces the four layers on a minimal example. This
vignette walks through the pieces a user needs once they start
depending on `genproc` for real work: the shape of the result, how
errors are reported, how the optional layers compose, and what the
current edges are.

```{r load}
library(genproc)
```

## A small working example

The same synthetic file-conversion task as in the README — one row
per file, `convert()` is the per-case function.

```{r workspace}
src_dir <- file.path(tempdir(), "genproc-vignette-src")
dst_dir <- file.path(tempdir(), "genproc-vignette-dst")
dir.create(src_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(dst_dir, showWarnings = FALSE, recursive = TRUE)

write.csv(head(iris),       file.path(src_dir, "a.csv"), row.names = FALSE)
write.csv(head(mtcars),     file.path(src_dir, "b.csv"), row.names = FALSE)
write.csv(head(airquality), file.path(src_dir, "c.csv"), row.names = FALSE)

convert <- function(src_dir, src_file, dst_dir, dst_file) {
  df <- read.csv(file.path(src_dir, src_file))
  saveRDS(df, file.path(dst_dir, dst_file))
}

mask <- data.frame(
  src_dir  = src_dir,
  src_file = c("a.csv", "b.csv", "c.csv"),
  dst_dir  = dst_dir,
  dst_file = c("a.rds", "b.rds", "c.rds"),
  stringsAsFactors = FALSE
)

result <- genproc(convert, mask)
```

## Anatomy of `genproc_result`

`result` is an S3 list with a stable, documented set of fields:

```{r result-names}
class(result)
names(result)
```

- `log`: one row per case. Columns, in order: `case_id`, the mask
  parameter values (in the order they appear in the mask), then
  `success`, `error_message`, `traceback`, `duration_secs`.
- `reproducibility`: the environment snapshot captured at run
  start (see below).
- `n_success`, `n_error`: summary counts.
- `duration_total_secs`: total wall-clock time.
- `status`: `"done"` for a synchronous run, `"running"` or
  `"error"` for a non-blocking run before `await()`.

These fields are guaranteed stable across minor versions; new
fields may be added (e.g. `worker_id` for parallel runs), but
existing ones will never be removed or renamed.

### The log

Column order is designed for a human scanning a run:

```{r log-head}
result$log
```

`case_id` is stable and index-based (`case_0001`, `case_0002`, ...)
for now. A content-based variant is on the roadmap, for use cases
where rows of the mask can be reordered between runs.

### The reproducibility snapshot

```{r repro-str}
str(result$reproducibility, max.level = 1)
```

- `timestamp`: `POSIXct`, start time of the run.
- `r_version`, `platform`, `os`, `locale`, `timezone`: system info.
- `packages`: named character vector of package -> version, for
  every package attached or loaded via namespace at run start.
- `mask_snapshot`: the exact mask used (not a copy of a reference —
  the value).
- `parallel`: `NULL` for a sequential run, or a plain list mirroring
  the `parallel_spec()` used. Dropped class to make the snapshot
  portable to serialization formats.
- `nonblocking`: same pattern, for `nonblocking_spec()`.
- `inputs`: a fingerprint of every file the mask refers to, or
  `NULL` if input tracking was disabled (`track_inputs = FALSE`).
  See the next section.

The snapshot lives inside the result. You can compare two results
by comparing their `$reproducibility` slots directly.

### Input fingerprinting

`reproducibility$inputs` is the layer that protects against silent
drift of upstream files. It is captured at t0 of the run, alongside
the rest of the snapshot.

```{r inputs-shape}
str(result$reproducibility$inputs, max.level = 1)
```

- `method`: currently `"stat"`. Reserved for future extensions
  (e.g. `"md5"` for content hashing).
- `files`: a deduplicated table of every file referenced by the
  mask, with its size in bytes and last-modified time. One row per
  unique path: a config file shared across 100 cases produces a
  single row.
- `refs`: the (`case_id`, `column`, `path`) triples saying who
  referenced what. Joins back to `files` by `path`.

#### Heuristic detection

By default, every character column of the mask whose non-NA values
are existing files (and contain a path separator) is treated as an
input column. The `mask` used in this vignette has its paths split
across `src_dir` and `src_file`, so the heuristic finds nothing
useful — `src_dir` is a directory (excluded), `src_file` values
are bare names (no separator). For a mask that holds absolute
paths directly:

```{r inputs-auto}
mask_paths <- data.frame(
  csv_in = file.path(src_dir, c("a.csv", "b.csv", "c.csv")),
  stringsAsFactors = FALSE
)
do_one <- function(csv_in) nrow(read.csv(csv_in))

run0 <- genproc(do_one, mask_paths)
run0$reproducibility$inputs$files
```

#### Overrides

- `genproc(..., input_cols = c("col1", "col2"))` bypasses the
  heuristic and tracks exactly the named columns. Paths that don't
  exist at capture time are recorded with `NA` size/mtime and a
  warning is emitted.
- `genproc(..., skip_input_cols = c("col"))` keeps the heuristic
  but excludes a column (useful when a label column happens to
  match an existing file in cwd).
- `genproc(..., track_inputs = FALSE)` disables tracking entirely.
  `result$reproducibility$inputs` is `NULL`.

`input_cols` and `skip_input_cols` are mutually exclusive. Mixing
them raises an error — the two flags express contradictory
intentions and the call should clarify.

#### Comparing runs with `diff_inputs()`

```{r inputs-diff}
# Rewrite a.csv with strictly more content (size changes)
write.csv(iris, file.path(src_dir, "a.csv"), row.names = FALSE)

run1 <- genproc(do_one, mask_paths)
diff_inputs(run0, run1)
```

`diff_inputs()` returns an S3 object (`genproc_input_diff`) with a
print method for human reading and named list components for
programmatic access (`$changed`, `$unchanged`, `$removed`, `$added`).
Files are matched by canonical absolute path; cross-machine
comparison would need a separate matcher and is out of scope for
this version.

`diff_inputs()` refuses to compare snapshots produced with
different `method`s (forward-compatible with a future hash mode).

## How errors are reported

A case that throws does **not** stop the run. Here we delete a
source file between two runs:

```{r broken}
file.remove(file.path(src_dir, "b.csv"))
result_broken <- genproc(convert, mask)

result_broken$n_success
result_broken$n_error
```

The failing row carries the error message and a filtered traceback:

```{r broken-row}
bad <- result_broken$log[!result_broken$log$success, ]
bad$error_message
cat(bad$traceback[1], "\n")
```

The traceback is captured via `withCallingHandlers()` at the moment
the error is thrown — it is the real R call stack, not a string
pulled out of `conditionMessage()`. The internal `tryCatch()` and
`withCallingHandlers()` frames are filtered out so the trace reads
like a normal R error.

Restore the file for subsequent sections:

```{r restore}
write.csv(head(mtcars), file.path(src_dir, "b.csv"), row.names = FALSE)
```

## Parameter renaming with `f_mapping`

If the function you already have uses parameter names that don't
match your mask's column names, `f_mapping` renames them in place
without touching the source:

```{r fmapping, eval = FALSE}
# `f` uses generic names; the mask uses domain names.
f <- function(input_dir, input_file, output_dir, output_file) {
  df <- read.csv(file.path(input_dir, input_file))
  saveRDS(df, file.path(output_dir, output_file))
}

genproc(
  f = f,
  mask = mask,
  f_mapping = c(
    "input_dir"   = "src_dir",
    "input_file"  = "src_file",
    "output_dir"  = "dst_dir",
    "output_file" = "dst_file"
  )
)
```

## Parallel execution in depth

`parallel_spec()` records intent only; the actual workers are
started lazily by `future` when the plan is resolved.

### Power-user pattern: manage the plan yourself

Across many `genproc()` calls in the same session, installing the
plan once amortizes worker startup:

```{r parallel-plan, eval = FALSE}
future::plan(future::multisession, workers = 6)

result_1 <- genproc(convert, mask, parallel = parallel_spec())
result_2 <- genproc(convert, mask, parallel = parallel_spec())
# reuses the same workers
```

### One-off pattern: let genproc install the plan

When `workers` is passed without `strategy`, `genproc()`
auto-defaults to `multisession` for that single call and restores
the previous plan on exit. This avoids the silent trap of
`workers = N` being ignored because the current plan is
`sequential`:

```{r parallel-oneoff, eval = FALSE}
genproc(convert, mask, parallel = parallel_spec(workers = 4))
```

### Amortization threshold

`multisession` costs roughly 1–3 seconds per worker at first use.
Below ~10 seconds of real work, a parallel run can be slower than
a sequential one. Don't read a single short benchmark as evidence
of a bug: run at workload size.

### RNG reproducibility

`parallel_spec(seed = TRUE)` (the default) draws independent
L'Ecuyer-CMRG streams from a random master seed. Identical master
seed — identical per-case RNG state regardless of worker count or
chunking. To pin the master, pass an integer:
`parallel_spec(seed = 42L)`.

## Non-blocking execution in depth

```{r nb-example, eval = FALSE}
job <- genproc(convert, mask, nonblocking = nonblocking_spec())
status(job)         # "running" or "done"
job <- await(job)   # blocks until resolution
job$log
```

### What's in the skeleton

Before `await()`:

- `log`, `n_success`, `n_error`, `duration_total_secs` are `NULL`.
- `reproducibility` is already populated — the snapshot is captured
  synchronously, before the future is submitted, so it reflects t0.
- `status` is `"running"`.
- The future itself is stored in `attr(x, "future")`.

After `await()`:

- All fields populated as in a synchronous run.
- `attr(x, "future")` is removed.
- `status` is `"done"` — or `"error"` if the wrapper future itself
  crashed (a rare case: per-case errors are caught by the logging
  layer and don't propagate to the wrapper).

### Idempotence

`await()` is idempotent. Calling it on an object that has already
been materialized (or was synchronous to begin with) returns it
unchanged. This makes it safe to pepper in user code without
tracking whether a particular result has already been collected.

### Default strategy: why `"multisession"`

Unlike `parallel_spec()`, `nonblocking_spec()` defaults to
`strategy = "multisession"`, not `NULL`. Rationale: a function
named "non-blocking" must not silently block because the current
`future::plan()` is `sequential` (the default in a fresh R
session). Power-users who manage their plan can pass
`strategy = NULL` explicitly to defer.

### Plan lifetime

When `nonblocking_spec()` installs a plan, the previous plan is
**not** restored on `genproc()` exit — it is restored by `await()`
once the future has been collected. Restoring earlier would call
`future::plan()` while the wrapper future is still running, which
shuts down the multisession workers and surfaces a "Future was
canceled" error at collection time. The trade-off is that if you
never call `await()`, the installed plan stays active for the rest
of the session. Power-users who pass `strategy = NULL` and manage
the plan themselves are not affected.

## Composition: parallel × non-blocking

The two optional layers are orthogonal:

```{r compose, eval = FALSE}
job <- genproc(
  convert, mask,
  parallel    = parallel_spec(workers = 6),
  nonblocking = nonblocking_spec()
)
# get control back immediately
# ... do other work ...
job <- await(job)
```

The non-blocking layer launches one outer future. Inside it, the
parallel layer dispatches cases via `future.apply`. Note: with both
set to `multisession`, `future.apply` detects it is already inside
a future and degrades the inner layer to `sequential` by default,
to avoid worker explosion. For true nested parallelism, install
`future::plan(list(...))` explicitly and pass `strategy = NULL` on
both specs.

## Current edges and roadmap

Not yet in the package, but explicitly planned:

- **Content-hash input fingerprinting**: the current `inputs` layer
  uses a stat-based fingerprint (size + mtime), which detects every
  legitimate file modification but can be fooled by an adversary who
  preserves both. A `method = "md5"` (or `"xxhash64"`) opt-in is
  reserved in the API and will land later.
- **Content-based `case_id`**: today case IDs are index-based. A
  content-based variant will make replay stable even if mask rows
  are reordered.
- **Error replay**: `replay(result, case_id)` to rerun one failed
  case in isolation.
- **Monitored progress**: opt-in progress reporting that survives
  non-blocking and parallel modes.
- **`cancel()` for non-blocking**: backend-dependent, deferred.

The architecture is designed so that adding these layers does not
require changes to existing user code — new layers are composed as
extra arguments to `genproc()`, and extra fields on
`genproc_result` accumulate without removing existing ones.
