---
title: "genproc: deep dive"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{genproc: deep dive}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

The `README` introduces the four layers on a minimal example. This
vignette walks through the pieces a user needs once they start
depending on `genproc` for real work: the shape of the result, how
errors are reported, how the optional layers compose, and what the
current edges are.

```{r load}
library(genproc)
```

## A small working example

The same synthetic file-conversion task as in the README — one row
per file, `convert()` is the per-case function.

```{r workspace}
src_dir <- file.path(tempdir(), "genproc-vignette-src")
dst_dir <- file.path(tempdir(), "genproc-vignette-dst")
dir.create(src_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(dst_dir, showWarnings = FALSE, recursive = TRUE)

write.csv(head(iris),       file.path(src_dir, "a.csv"), row.names = FALSE)
write.csv(head(mtcars),     file.path(src_dir, "b.csv"), row.names = FALSE)
write.csv(head(airquality), file.path(src_dir, "c.csv"), row.names = FALSE)

convert <- function(src_dir, src_file, dst_dir, dst_file) {
  df <- read.csv(file.path(src_dir, src_file))
  saveRDS(df, file.path(dst_dir, dst_file))
}

mask <- data.frame(
  src_dir  = src_dir,
  src_file = c("a.csv", "b.csv", "c.csv"),
  dst_dir  = dst_dir,
  dst_file = c("a.rds", "b.rds", "c.rds"),
  stringsAsFactors = FALSE
)

result <- genproc(convert, mask)
```

## Anatomy of `genproc_result`

`result` is an S3 list with a stable, documented set of fields:

```{r result-names}
class(result)
names(result)
```

- `log`: one row per case. Columns, in order: `case_id`, the mask
  parameter values (in the order they appear in the mask), then
  `success`, `error_message`, `traceback`, `duration_secs`.
- `reproducibility`: the environment snapshot captured at run
  start (see below).
- `n_success`, `n_error`: summary counts.
- `duration_total_secs`: total wall-clock time.
- `status`: `"done"` for a synchronous run, `"running"` or
  `"error"` for a non-blocking run before `await()`.

These fields are guaranteed stable across minor versions; new
fields may be added (e.g. `worker_id` for parallel runs), but
existing ones will never be removed or renamed.

### The log

Column order is designed for a human scanning a run:

```{r log-head}
result$log
```

`case_id` is stable and index-based (`case_0001`, `case_0002`, ...)
for now. A content-based variant is on the roadmap, for use cases
where rows of the mask can be reordered between runs.

### The reproducibility snapshot

```{r repro-str}
str(result$reproducibility, max.level = 1)
```

- `timestamp`: `POSIXct`, start time of the run.
- `r_version`, `platform`, `os`, `locale`, `timezone`: system info.
- `packages`: named character vector of package -> version, for
  every package attached or loaded via namespace at run start.
- `mask_snapshot`: the exact mask used (not a copy of a reference —
  the value).
- `parallel`: `NULL` for a sequential run, or a plain list mirroring
  the `parallel_spec()` used. Dropped class to make the snapshot
  portable to serialization formats.
- `nonblocking`: same pattern, for `nonblocking_spec()`.

The snapshot lives inside the result. You can compare two results
by comparing their `$reproducibility` slots directly.

## How errors are reported

A case that throws does **not** stop the run. Here we delete a
source file between two runs:

```{r broken}
file.remove(file.path(src_dir, "b.csv"))
result_broken <- genproc(convert, mask)

result_broken$n_success
result_broken$n_error
```

The failing row carries the error message and a filtered traceback:

```{r broken-row}
bad <- result_broken$log[!result_broken$log$success, ]
bad$error_message
cat(bad$traceback[1], "\n")
```

The traceback is captured via `withCallingHandlers()` at the moment
the error is thrown — it is the real R call stack, not a string
pulled out of `conditionMessage()`. The internal `tryCatch()` and
`withCallingHandlers()` frames are filtered out so the trace reads
like a normal R error.

Restore the file for subsequent sections:

```{r restore}
write.csv(head(mtcars), file.path(src_dir, "b.csv"), row.names = FALSE)
```

## Parameter renaming with `f_mapping`

If the function you already have uses parameter names that don't
match your mask's column names, `f_mapping` renames them in place
without touching the source:

```{r fmapping, eval = FALSE}
# `f` uses generic names; the mask uses domain names.
f <- function(input_dir, input_file, output_dir, output_file) {
  df <- read.csv(file.path(input_dir, input_file))
  saveRDS(df, file.path(output_dir, output_file))
}

genproc(
  f = f,
  mask = mask,
  f_mapping = c(
    "input_dir"   = "src_dir",
    "input_file"  = "src_file",
    "output_dir"  = "dst_dir",
    "output_file" = "dst_file"
  )
)
```

## Parallel execution in depth

`parallel_spec()` records intent only; the actual workers are
started lazily by `future` when the plan is resolved.

### Power-user pattern: manage the plan yourself

Across many `genproc()` calls in the same session, installing the
plan once amortizes worker startup:

```{r parallel-plan, eval = FALSE}
future::plan(future::multisession, workers = 6)

result_1 <- genproc(convert, mask, parallel = parallel_spec())
result_2 <- genproc(convert, mask, parallel = parallel_spec())
# reuses the same workers
```

### One-off pattern: let genproc install the plan

When `workers` is passed without `strategy`, `genproc()`
auto-defaults to `multisession` for that single call and restores
the previous plan on exit. This avoids the silent trap of
`workers = N` being ignored because the current plan is
`sequential`:

```{r parallel-oneoff, eval = FALSE}
genproc(convert, mask, parallel = parallel_spec(workers = 4))
```

### Amortization threshold

`multisession` costs roughly 1–3 seconds per worker at first use.
Below ~10 seconds of real work, a parallel run can be slower than
a sequential one. Don't read a single short benchmark as evidence
of a bug: run at workload size.

### RNG reproducibility

`parallel_spec(seed = TRUE)` (the default) draws independent
L'Ecuyer-CMRG streams from a random master seed. Identical master
seed — identical per-case RNG state regardless of worker count or
chunking. To pin the master, pass an integer:
`parallel_spec(seed = 42L)`.

## Non-blocking execution in depth

```{r nb-example, eval = FALSE}
job <- genproc(convert, mask, nonblocking = nonblocking_spec())
status(job)         # "running" or "done"
job <- await(job)   # blocks until resolution
job$log
```

### What's in the skeleton

Before `await()`:

- `log`, `n_success`, `n_error`, `duration_total_secs` are `NULL`.
- `reproducibility` is already populated — the snapshot is captured
  synchronously, before the future is submitted, so it reflects t0.
- `status` is `"running"`.
- The future itself is stored in `attr(x, "future")`.

After `await()`:

- All fields populated as in a synchronous run.
- `attr(x, "future")` is removed.
- `status` is `"done"` — or `"error"` if the wrapper future itself
  crashed (a rare case: per-case errors are caught by the logging
  layer and don't propagate to the wrapper).

### Idempotence

`await()` is idempotent. Calling it on an object that has already
been materialized (or was synchronous to begin with) returns it
unchanged. This makes it safe to pepper in user code without
tracking whether a particular result has already been collected.

### Default strategy: why `"multisession"`

Unlike `parallel_spec()`, `nonblocking_spec()` defaults to
`strategy = "multisession"`, not `NULL`. Rationale: a function
named "non-blocking" must not silently block because the current
`future::plan()` is `sequential` (the default in a fresh R
session). Power-users who manage their plan can pass
`strategy = NULL` explicitly to defer.

### Plan lifetime

When `nonblocking_spec()` installs a plan, the previous plan is
**not** restored on `genproc()` exit — it is restored by `await()`
once the future has been collected. Restoring earlier would call
`future::plan()` while the wrapper future is still running, which
shuts down the multisession workers and surfaces a "Future was
canceled" error at collection time. The trade-off is that if you
never call `await()`, the installed plan stays active for the rest
of the session. Power-users who pass `strategy = NULL` and manage
the plan themselves are not affected.

## Composition: parallel × non-blocking

The two optional layers are orthogonal:

```{r compose, eval = FALSE}
job <- genproc(
  convert, mask,
  parallel    = parallel_spec(workers = 6),
  nonblocking = nonblocking_spec()
)
# get control back immediately
# ... do other work ...
job <- await(job)
```

The non-blocking layer launches one outer future. Inside it, the
parallel layer dispatches cases via `future.apply`. Note: with both
set to `multisession`, `future.apply` detects it is already inside
a future and degrades the inner layer to `sequential` by default,
to avoid worker explosion. For true nested parallelism, install
`future::plan(list(...))` explicitly and pass `strategy = NULL` on
both specs.

## Current edges and roadmap

Not yet in the package, but explicitly planned:

- **Input file hashing**: the `reproducibility` layer records the
  mask, but does not yet hash file inputs referenced in the mask.
  When a file changes without its path changing, that drift is
  currently invisible — this will be flagged.
- **Content-based `case_id`**: today case IDs are index-based. A
  content-based variant will make replay stable even if mask rows
  are reordered.
- **Error replay**: `replay(result, case_id)` to rerun one failed
  case in isolation.
- **Monitored progress**: opt-in progress reporting that survives
  non-blocking and parallel modes.
- **`cancel()` for non-blocking**: backend-dependent, deferred.

The architecture is designed so that adding these layers does not
require changes to existing user code — new layers are composed as
extra arguments to `genproc()`, and extra fields on
`genproc_result` accumulate without removing existing ones.
